Galaxy image statistical generator: how to be confident?

### 2024-09-04 paper/main.tex
more on intro
starts generator descriptions
VAE

\documentclass[11pt]{amsart}
\usepackage{amsaddr}
\usepackage{mathtools}
\usepackage{aas_macros} % macro pour Bibtex
\usepackage{hyperref}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsthm}
\usepackage{amsmath}
% \usepackage{makecell}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{color}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{natbib,fancyhdr} %new

@@

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.

@@

\section{Summary}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{elsarticle-harv}%{aa}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass[11pt]{amsart}
\usepackage{amsaddr}
\usepackage{mathtools}
\usepackage{aas_macros} % macro pour Bibtex
\usepackage{hyperref}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsthm}
\usepackage{amsmath}
% \usepackage{makecell}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{color}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{natbib,fancyhdr} %new
\usepackage{bm}

@@

\newcommand{\Esp}[0]{\ensuremath{\mathbb{E}}}
\newcommand{\DKL}[0]{\ensuremath{\mathbb{D}_{\mathsf{KL}}}}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.
%
\section{Generative Models}
%
The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ (\textit{pdf}) and aim to give a representation and leads to different strategies.
%
\subsection{Variational Auto-Encoder}
%
Using \texttt{VAE} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (\textit{latent variable}) whose \textit{a priori} known $\pi(\bm{z})$ \textit{pdf} such that $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$. Although, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ and the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $n$ data samples ($\{\bm{x}^{i}\}_{i\leq n}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 
$\int\ p_{\bm{\theta}}(\bm{x}|\bm{z})\pi(\bm{z}) dz$. To overcome such problem, one uses first the Bayes relation
\begin{equation}
p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}}(\bm{x}|\bm{z}) \pi(\bm{z})
\end{equation}
and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$. If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{x\sim p}[\log(p(x)/q(x))]$, then taking the expected value according to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
\begin{align}
\log p(\bm{x}) = \log p_{\bm{\theta},\bm{\phi}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
\end{align}
yields 
\begin{align}
\log p(\bm{x})  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
&\geq \matcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
\end{align}
where we have emphasize the \textit{evidence lower bound} (aka ELBO)
\begin{equation}
\matcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z}))
\end{equation}
and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$.

It is usual to call $p_{\bm{\theta}}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. Once the optimization is done, then decoder is used as a $\bm{x}$ sample generator when it is fed with $\bm{z}$ samples according to $\bm{z}\sim\pi(\bm{z})$.

%
\section{Experiment}

@@

\section{Discussion}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{elsarticle-harv}%{aa}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Details on the dataset}
